Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-26 paper.tex

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-Cov-2 ($n$=3,960,704) % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and GPUs.

@@

\subsection{Extensibility: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}
to an AWS S3 standard bucket. 
We repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

@@

\subsection{Extensibility: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging the
massively parallel architecture of GPUs to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
% (Nickolls & Dally, 2010;
% Mittal & Vetter, 2015; 
% Owens et al., 2008). 
This has profound implications for
fields ranging from molecular dynamics~\citep{stone2007accelerating}
% (Stone et al., 2010) 
to deep learning~\citep{lecun2015deep}
% (LeCun et al., 2015). 
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
%(Okuta et al., 2017), 
and
PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
% (Klöckner, 2012) 
has democratized access to GPU acceleration.
These tools offer user-friendly interfaces and just-in-time compilation,
allowing scientists to harness the power of GPUs within the familiar Python
ecosystem, thereby significantly accelerating scientific discovery and
broadening the adoption of accelerated computing across various research
domains.

@@

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data in a large ecosystem of interoperating tools.
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the present generation
of bioinformatics tools. Large scale VCF data cannot
currently be
processed without incurring a substantial economic
(and environmental~\cite{grealey2022carbon}) cost.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide an efficient and scalable
tool to perform conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly,
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends
quadratically on the number of alleles (although practical
solutions have been suggested that we plan to
implement~\cite{poterba2024scalable}).
Secondly, the design of
VCF Zarr emphasises efficiency of analysis for a fixed
dataset, and does not consider how samples (and the
corresponding novel variants) should be added.
Thirdly, Zarr works best for numerical data of a fixed
dimension, and therefore may not suitable for representing
the unstructured data often included in VCF INFO fields.

@@

[TODO REWRITE]
The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed
to provide a viable alternative to standard bioinformatics workflows.
Two initial directions for development, however, may quickly
yield sufficient results to both greatly improve researcher productivity on
large, centrally managed datasets such as Genomics England
and motivate further research and development.
The first direction is to provide compatibility with existing
workflows via a ``\texttt{vcztools}'' command line utility which
implements a subset of \texttt{bcftools} functionality (such as
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders
of magnitude, and reduce the need for user orchestration of
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

[TODO REWRITE]
The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy}
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters.
This scaling is achieved by distributing calculations over
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created
to facilitate work on a scalable genetics toolkit for
Python~\cite{sgkit2024} built on Xarray.
While the high-level facilities for distributed
computation provided by Xarray
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here
illustrate that working at the lowest level,
by sequentially applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.

@@

\subsection{vcztools}
Vcztools is available for download on the Python package index, and
supports all major platforms [TODO do we support Windows??],
and is written in a combination of Python and C.
We use
the Zarr-Python package~\citep{zarrpython} for all Zarr operations.
The mini-language in the \texttt{query} command is parsed using
PyParsing [TODO cite].
As a general principle, data is only retrieved from the Zarr store
when it is required to fulfil a particular query. For example,
when a subset of samples are specified, we only retrieve chunks
in arrays such as \texttt{call\_genotype} for those
samples. Chunks in the variants dimension are retrieved as needed,
and we use a multithreaded strategy to fetch and decode variant chunks
while another thread encodes VCF. Generating VCF text at speed
is a challenging problem, and we use a C extension module to convert
a given set of Zarr chunks represented as numpy arrays to VCF.
Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
search queries.

>>>>

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-Cov-2 ($n$=3,960,704), % TODO update for tranche 2 data 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and GPUs.

@@

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}
to an AWS S3 standard bucket. 
We repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

@@

\subsection{Future applications: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging the
massively parallel architecture of GPUs to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
% (Nickolls & Dally, 2010;
% Mittal & Vetter, 2015; 
% Owens et al., 2008). 
This has profound implications for
fields ranging from molecular dynamics~\citep{stone2007accelerating}
% (Stone et al., 2010) 
to deep learning~\citep{lecun2015deep}
% (LeCun et al., 2015). 
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
%(Okuta et al., 2017), 
and
PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
% (Klöckner, 2012) 
has democratized access to GPU acceleration.
These tools offer user-friendly interfaces and just-in-time compilation,
allowing scientists to harness the power of GPUs within the familiar Python
ecosystem, thereby significantly accelerating scientific discovery and
broadening the adoption of accelerated computing across various research
domains.

@@

\section{Discussion}
% Zarr is great
VCF is a central element of modern genomics, facilitating
the exchange of data in a large ecosystem of interoperating tools.
Its current row-oriented form, however,
is fundamentally inefficient,
profoundly limiting the scalability of the present generation
of bioinformatics tools. Large scale VCF data cannot
currently be
processed without incurring a substantial economic
(and environmental~\cite{grealey2022carbon}) cost.
We have shown here that this is not a necessary situation,
and that greatly improved efficiency can be achieved by
using more appropriate storage representations tuned
to the realities of modern computing. We have argued that
Zarr provides a powerful basis for cloud-based
storage and analysis of large-scale genetic variation data.
We propose the VCF Zarr specification which losslessly
maps VCF data to Zarr, and provide the basic software
infrastructure required for bidirectional conversion.

% We're not pretending is a solution to all problems and
% up front about weaknesses of the model.
Zarr provides pragmatic solutions to some of the more pressing
problems facing the analysis of large-scale genetic variation
data, but it is not a panacea. Firstly,
any dataset containing a variant with a large number of alleles
(perhaps due to indels) will cause problems because the
dimensions of fields are determined by their \emph{maximum}
dimension among all variants. In particular this is problematic
for fields like PL in which the dimension depends
quadratically on the number of alleles.
While the local alleles representation~\cite{poterba2024scalable}
is a substantial improvement (see All of Us and Norway Spruce
case studies) it may be lossy (see Methods). 
Secondly, the design of
VCF Zarr emphasises efficiency of analysis for a fixed
dataset, and does not consider how samples (and the
corresponding novel variants) should be added.
Thirdly, Zarr works best for numerical data of a fixed
dimension, and therefore may not suitable for representing
the unstructured data often included in VCF INFO fields.

@@

% Streaming is inherently incompatible with efficient utilisation
% of cloud object stores.
Modern computing is increasingly parallel, from multicore CPUs, GPUs 
and the massively parallel scale provided by public clouds.
Making optimal use of these parallel resources requires new programming
models, and in particular adapting analytics workloads to the properties of 
disaggregated cloud object stores is essential.
Classical workflows in genomics emphasise streaming, 
in which large volumes of data are read sequentially from file,
streamed through the processor, and the results written sequentially to 
storage or to the input of another process in a pipeline. 
Using this model on data held on a cloud object store requires
us to emulate file-system semantics,
and places hard requirements on receiving data in a specific order.
As Durner et al~\cite{durner2023exploiting} show (and we confirm 
in Fig~\ref{fig-s3-network-throughput}), sustained data throughput speeds similar to
a locally attached SSD can be obtained from cost-effective and 
massively scalable cloud object stores like Amazon S3,
but on the condition that we do \emph{not} place constraints on
the order in which data is received (due to the high variance in latency).
In this context, Zarr provides a simple and elegant model for computation
in which the \emph{chunk} is the basic unit of parallelism. For many
applications, each chunk can be processed independently and in no particular order.

% In the streaming model, large volumes of data are read sequentially from storage,
% streamed through the processor, with results written sequentially to 
% storage or read by other processes in a pipeline. 
% Row-oriented storage of data (see the Introduction) is also well-suited to 
% streaming computation
% % The focus on large sequential 
% % reads is derived ultimately from the realities of disk drive storage, and leads
% % to efficient utilisation through multiple layers of operating system and 
% % on-device caching, with heuristic parameters tuned through decades of experience.
% % While there is some potential for parallelism in chaining multiple processes
% % in a Unix pipeline, it tends to be limited by the slowest step and is not
% % an effective way to use the tens of CPU cores available on a modern machines.
% Splitting work up along the genome has provided ample opportunities to 
% to parallelise in the past, and is part of the motivation 
% for the current practise of providing VCF data in sub-chromosome partitions 
% (see the Genomics England, Our Future Health and Norway Spruce case-studies). 
% However, with increasingly large sample sizes, this approach to parallelism

There is therefore an opportunity to develop a new 
generation of Zarr-native applications, taking advantage of the efficient data
representation and natural model of chunk-based parallelism.
Although there are several independent Zarr implementations
across programming languages (see Methods),
the Python data science ecosystem is a particularly powerful
platform,
with a rich suite of tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy}
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters.
This scaling is achieved by distributing calculations over
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created
to facilitate work on a scalable genetics toolkit for
Python~\cite{sgkit2024} built on Xarray.
While the high-level facilities for distributed
computation provided by Xarray
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here
illustrate that working at the lowest level,
by applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.

@@

\subsection{vcztools}
Vcztools is available for download on the Python package index, and
supports all major platforms [TODO do we support Windows??],
and is written in a combination of Python and C.
We use
the Zarr-Python package~\citep{zarrpython} for all Zarr operations.
The mini-language in the \texttt{query} command is parsed using
PyParsing [TODO cite].
[Add a note here on why it was easier to to write a Python application
from scratch than to update bcftools.]

As a general principle, data is only retrieved from the Zarr store
when it is required to fulfil a particular query. For example,
when a subset of samples are specified, we only retrieve chunks
in arrays such as \texttt{call\_genotype} for those
samples. Chunks in the variants dimension are retrieved as needed,
and we use a multithreaded strategy to fetch and decode variant chunks
while another thread encodes VCF. Generating VCF text at speed
is a challenging problem, and we use a C extension module to convert
a given set of Zarr chunks represented as numpy arrays to VCF.
Based on the \texttt{region\_index} array computed by \texttt{vcf2zarr},
we use PyRanges~\citep{stovner2020pyranges} to perform efficient interval
search queries.
